Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-01 doc/papers/cgo2014/paper.tex
CGO: review pages 1-3

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However the
    computation kernels often differ from the classical ones as these languages
    favor a high-level coding style, where loops are often replaced by calls
    to the relevant toolbox/module/library/package. As a consequence, traditional
    loop optimization techniques like loop fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

@@

The Python language~\cite{rossum97} now provides a full toolkit to carry out
scientific experiments~\cite{scipy}, from core scientific routines with the
Numpy package\cite{oliphant2007,numpyarray2011}, to scientific packages with
the Scipy package, plotting facilities with the Matplotlib package, enhanced
terminal and notebooks with IPython. As a consequence, there has been a move
from historical languages like Fortran to Python, as showcased by the success
of the Scipy conference.

@@

A query using the terms \emph{numpy} and \emph{slow} yields many results, many
of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
``A: rewrite it that way'', where the original code performs explicit looping
over an array, and the rewritten code uses the relevant combination of numpy
function.

For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
roughly achieves a $times 4$ speedup over the loop version.

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, label={lst:l2norm-loopy}]
def slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, label={lst:l2norm}]
import numpy as np
def l2norm(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

Our guess is that beginner Numpy users are more likely to make use of raw loops
similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
API and of the good programming practice grows, they get to write code similar
to Listing~\ref{lst:l2norm}. As a direct consequence, a compiler should
focus on optimizing high-level constructs rather than raw loops. Both to encourage good
practice with respect to Numpy code, and because this is the kind of code Numpy
users ultimately write.

@@

\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

@@

In Numpy, all point-to-point array operations allocate a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage and add an extra
allocation/disallocation costly operation. In the rosen expression from
Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
create a temporary array but a view) to hold intermediate steps. Had the
computation been lazy, no temporary would have been needed.

@@

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. With a smamm set of functions, a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed with operator fusion.

@@

Even if compiled with vectorization flags on, there would be very little
benefit to generate SIMD instructions for the respective array operations used
by each operator, as the memory load store would have dominated, especially as
Numpy typically operates on double precision floats. In a similar manner,
parallelization of each operator would yield a terrible speedup because a
synchronization fence would need to be issued between each temporary
computation and the loop computation intensity would be very low compared to
the memory pressure implied by two array reads and one array write for a single
binary operator.

At the opposite, if all computation were merged into a single loop using
Temporaries Elimination and Operator fusion as described above, parallelization
would be applicable as barrier between binary operation are removed and only
one load and one store will be use for multiple operations.

@@

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}~\cite{numba}
compiler uses additional type information to generate sequential LLVM bytecode.
Numba uses Just-In-Time(JIT) compilation and do not suffer from
backward-incompatibility issues. When meeting an unsupported construct, numba
falls back to Python C-API calls.

@@

Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is
Python to LLVM bytecode translation, but limits its scope to the Numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between Numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s. As it applies optimizations like operator fusion, it
encourages programmer to express their code using the higher-level construct
available with Numpy.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code. As it requires to write
functional code, it needs code modifications of \textit{Pythonic} code

@@

Thanks to variadic templates introduced in C++11, it is possible to write a
generic C++ meta-class that implements the same API as numpy array and provides
all unary and binary operators, as well as the various form of indexing and
slicing. Listing~\ref{lst:rosen-cxx} illustrates the translation of the Python
code from Listing~\ref{lst:rosen} into C++. Although this version is more
verbose than the Python one, the translation from one to the other does not
raise any particular issue.

@@

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
typedef typename x::dtype T;
constexpr size_t N = x::ndim;
ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

The computation of read-write effects involve the annotation of numpy functions
so that the Python compiler knows that most of them do not modify their
arguments, and the computation of argument effects for user functions. This can
be done by considering each function individually and looking for the two kind
of constructs that can modify a function argument: indexing and further
function calls.

@@

A first approach would be to use recursive tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}
% Les taches sont en général plus lente que les boucles for (malheureusement)
% et il y a des fois ou on ne pourra jamais être plus rapide si on passe par
% des taches.
% Souvent, c'est a cause du scheduling static qui permet d'optimiser le cache
% et aussi parce que les compilos, quand ils voient un omp for static, ils
% font directement du déroulage de boucle. (Jusqu'au full unrolling des fois
% avec autant de tache que de thread). 
%
% Dans notre cas, on ne connait pas forcement la taille de l'espace d'iteration
% a la compilation. On ne pourra donc pas faire trop de unrolling mais bon, on
% doit pouvoir faire comme gcc. Par contre, pour l'optimisation du cache, les
% boucles for assurent que c'est le même noeud qui exécute le même "chunk" entre
% les boucles for et ça, on le perd avec les taches. Je ne sais pas si ça joue
% vraiment.
% Autre problem. On va créé des taches en assez grande quantité. Ce que libgomp
% supporte assez mal. Aussi, on va créé du datasharing pour des petites taches
% (iteration de longueur 1 dans le cas du (1, 1000000)) ce qui est couteux et
% assez mal supporté par libgomp. Il faudra voir comment ça se passe avec libiomp
% qui est utilisé par clang.
%
% Au final : Ca doit être jouable. Il y a plusieurs raisons qui peuvent faire
% qu'on a de mauvaise perfs mais sans mesures, on ne sait pas.

@@

\subsection{Performance of the rosenbrock Function}

>>>>

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However as 
    these languages favor a high-level coding style, the computation kernels are
    different from what we use to see in traditional C or Fortran programs:
    explicit raw loops are replaced by high-level constructs and calls to the 
    relevant toolbox/module/library/package. As a consequence, traditional
    loop optimization techniques like fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

@@

The Python language~\cite{rossum97} has a rich ecosystem that now provides a 
full toolkit to carry out scientific experiments, from core scientific routines 
with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific 
packages with the Scipy package~\cite{scipy}, plotting facilities with the 
Matplotlib package, enhanced terminal and notebooks with IPython. As a 
consequence, there has been a move from historical languages like Fortran to 
Python, as showcased by the success of the Scipy conference.

@@

A query using the terms \emph{numpy} and \emph{slow} yields many results, many
of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
``A: rewrite it that way'', where the original code performs explicit looping
over an array, and the rewritten code uses the relevant combination of Numpy
function.

For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
roughly achieves a $\times 4$ speedup over the loop version.

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
def slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in Numpy.}, label={lst:l2norm}]
import numpy as np
def l2norm(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

Newcommers to Python with a prior C or Fortran experience are likely to make 
use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their 
knowledge of the Numpy API and of the good programming practice grows, they get to write code similar
to Listing~\ref{lst:l2norm}. As a direct consequence, 
\todo{Serge: don't you think we should also says that high-level constructs 
enable better optimizations for the compiler?}
a compiler should focus on optimizing high-level constructs rather than raw 
loops. Both to encourage good practice with respect to Numpy code, and because 
this is the kind of code Numpy users ultimately write.

@@

\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy}, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

@@

In Numpy, all point-to-point array operations allocate a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage and add an extra
allocation/disallocation costly operation. In the \textit{rosen} function from
Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
create a temporary array but a view) to hold intermediate steps. Had the
computation been lazy, no temporary would have been needed.

@@

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. The Listing~\ref{lst:rosen} 
illustrates the use of a small set of functions: a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed with operator fusion.

@@

Without operator fusion, there would be very little benefit to generate SIMD 
instructions for the respective array operations used by each operator, as the
memory load store would have dominated. This is even more important as
Numpy typically operates on double precision floats. Parallelization also 
suffers from the lack of operator fusion: a synchronization fence is needed
between each temporary computation. 
Hence the loop computation intensity would be very low compared to
the memory pressure implied by two array reads and one array write for a single
binary operator.

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barrier between binary operations are no longer needed and loads
and stores will be used to perform multiple operations.

@@

The numba 
compiler uses additional type information to generate sequential LLVM bytecode.
Numba uses Just-In-Time(JIT) compilation and do not suffer from
backward-incompatibility issues. When meeting an unsupported construct, numba
falls back to Python C-API calls.

@@

Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is
Python to LLVM bytecode translation, but limits its scope to the Numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between Numpy functions and a set
of parallel primitives including \texttt{map}, \texttt{scan}, and
\texttt{reduce}. As it applies optimizations such as operator fusion, it
encourages programmer to express their code using the higher-level construct
available with Numpy.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, Numpy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code. As it requires to write
functional code, it needs code modifications of \textit{Pythonic} code

@@

Thanks to variadic templates introduced in C++11, it is possible to write a
generic C++ meta-class that implements the same API as Numpy array and provides
all unary and binary operators, as well as the various form of indexing and
slicing. Listing~\ref{lst:rosen-cxx} illustrates the translation of the Python
code from Listing~\ref{lst:rosen} into C++. Although this version is more
verbose than the Python one, the translation from one to the other does not
raise any particular issue.

@@

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the Rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
typedef typename x::dtype T;
constexpr size_t N = x::ndim;
ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

The computation of read-write effects involve the annotation of Numpy functions
so that the Python compiler knows that most of them do not modify their
arguments, and the computation of argument effects for user functions. This can
be done by considering each function individually and looking for the two kind
of constructs that can modify a function argument: indexing and further
function calls.

@@

A first approach would be to use recursive tasks and manual loop chunking.
\todo{pierrick ? c'est bête comme idée ?}
% Les taches sont en général plus lente que les boucles for (malheureusement)
% et il y a des fois ou on ne pourra jamais être plus rapide si on passe par
% des taches.
% Souvent, c'est a cause du scheduling static qui permet d'optimiser le cache
% et aussi parce que les compilos, quand ils voient un omp for static, ils
% font directement du déroulage de boucle. (Jusqu'au full unrolling des fois
% avec autant de tache que de thread). 
%
% Dans notre cas, on ne connait pas forcement la taille de l'espace d'iteration
% a la compilation. On ne pourra donc pas faire trop de unrolling mais bon, on
% doit pouvoir faire comme gcc. Par contre, pour l'optimisation du cache, les
% boucles for assurent que c'est le même noeud qui exécute le même "chunk" entre
% les boucles for et ça, on le perd avec les taches. Je ne sais pas si ça joue
% vraiment.
% Autre problem. On va créé des taches en assez grande quantité. Ce que libgomp
% supporte assez mal. Aussi, on va créé du datasharing pour des petites taches
% (iteration de longueur 1 dans le cas du (1, 1000000)) ce qui est couteux et
% assez mal supporté par libgomp. Il faudra voir comment ça se passe avec libiomp
% qui est utilisé par clang.
%
% Au final : Ca doit être jouable. Il y a plusieurs raisons qui peuvent faire
% qu'on a de mauvaise perfs mais sans mesures, on ne sait pas.

@@

\subsection{Performance of the Rosenbrock Function}
